---
title: "BRM on the bike dataset (regression)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BRM on the bike dataset (regression)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
set.seed(1234)
```

## Overview

This vignette demonstrates Blockwise Reduced Modeling (BRM) on the
**Capital Bikeshare** dataset, a regression problem where we predict hourly
ride counts (`cnt`) from weather and calendar features. Because `bike` is
otherwise complete, we induce a realistic blockwise missing pattern with
`simulate_blockwise_missing()` to show BRM in its element.

The method, and this dataset's role as a benchmark, are described in
Srinivasan, Currim, and Ram (2025), *A Reduced Modeling Approach for Making
Predictions With Incomplete Data Having Blockwise Missing Patterns*, INFORMS
Journal on Data Science.

```{r load}
library(blockwise)
data(bike)
str(bike, list.len = 20)
```

## Inducing a blockwise missing pattern

We mask two groups of columns jointly — mimicking the situation where two
independent data sources feeding your pipeline fail on different subsets of
rows — plus a small per-column noise rate.

```{r mask}
bike_miss <- simulate_blockwise_missing(
  bike,
  blocks = list(
    c("windspeed", "hum", "weekday"),
    c("hr", "temp", "weathersit")
  ),
  prop_missing = 0.30,
  noise        = 0.05
)

round(colMeans(is.na(bike_miss)) * 100, 1)  # percent missing per column
```

## Train / test split

```{r split}
set.seed(1234)
idx <- sample(nrow(bike_miss), size = floor(0.75 * nrow(bike_miss)))
train <- bike_miss[idx, ]
test  <- bike_miss[-idx, ]

X_train <- train[, setdiff(names(train), "cnt")]
y_train <- train$cnt

X_test  <- test[, setdiff(names(test), "cnt")]
y_test  <- test$cnt
```

## Fit BRM

`brm()` is learner-agnostic: pass any `learner()` specification. Here we try
a linear model and a gradient-boosted tree ensemble. The number of blocks is
chosen automatically by the elbow heuristic (`choose_num_blocks`).

```{r fit-lm}
fit_lm <- brm(X_train, y_train, learner = learner_lm())
fit_lm
```

```{r fit-gbm, eval = requireNamespace("gbm", quietly = TRUE)}
fit_gbm <- brm(
  X_train, y_train,
  learner  = learner_gbm(distribution = "poisson", n.trees = 300),
  n_blocks = fit_lm$n_blocks     # reuse so models are comparable
)
fit_gbm
```

## Predict and score

```{r rmse}
rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))

pred_lm <- predict(fit_lm, X_test)
cat("BRM (lm)   RMSE:", round(rmse(y_test, pred_lm), 2), "\n")
```

```{r rmse-gbm, eval = requireNamespace("gbm", quietly = TRUE)}
pred_gbm <- predict(fit_gbm, X_test)
cat("BRM (gbm)  RMSE:", round(rmse(y_test, pred_gbm), 2), "\n")
```

## Comparison to a listwise-deletion baseline

The conventional alternative is to drop rows that have any missing value and
fit a single model on what remains. This wastes data and gets progressively
worse as the missing rate grows.

```{r listwise}
complete_train <- na.omit(train)
fit_lw <- lm(cnt ~ ., data = complete_train)

# For a fair comparison we need to impute the test set's NAs somehow; use
# mean/mode from the complete training rows.
X_test_imp <- X_test
for (j in names(X_test_imp)) {
  na_idx <- is.na(X_test_imp[[j]])
  if (!any(na_idx)) next
  ref <- complete_train[[j]]
  if (is.factor(ref)) {
    X_test_imp[[j]][na_idx] <- names(sort(table(ref), decreasing = TRUE))[1]
  } else {
    X_test_imp[[j]][na_idx] <- mean(ref, na.rm = TRUE)
  }
}
pred_lw <- predict(fit_lw, newdata = X_test_imp)

cat("Listwise-deletion lm  RMSE:", round(rmse(y_test, pred_lw), 2), "\n")
cat("Training rows used    : BRM =", nrow(train), " listwise =",
    nrow(complete_train), "\n")
```

BRM keeps all training rows (splitting them into per-pattern subsets); the
listwise baseline throws away any row with at least one `NA`.

## Citation

If you use BRM in your work, please cite:

> Srinivasan, K., Currim, F., and Ram, S. (2025). *A Reduced Modeling
> Approach for Making Predictions With Incomplete Data Having Blockwise
> Missing Patterns.* INFORMS Journal on Data Science.

Run `citation("blockwise")` to get a ready-to-paste BibTeX entry.
